Information-theoretic measures for non-linear causality detection: \\ application to social media sentiment and cryptocurrency prices

### 2020-07-25 paper/submission.tex
Checkpoint with all revisions made and highlighted

  In modern portfolio theory, investors commonly calculate correlations between asset types to construct portfolios aiming to maximise their return at a given level of risk \cite{markowitz1952portfolio}. In the search for excess returns, quantitative approaches are often exploited to detect predictive signals across time series. {\color{blue} This may be considered as a field of price discovery, which represents a large and fundamental area of financial theory.} In the ideal case {with complete predictive information, \color{blue}from} knowing the movement of one price, the movement of a second can be inferred {\color{blue} exactly. In reality information is incomplete and held asymmetrically, and the ability to reliably obtain this and contribute to price discovery is often well rewarded}. In this paper we explore the effectiveness of two promising techniques for detecting information transfer between time series, applying these to identify causal signals between alternative data and cryptocurrency {\color{blue} price returns}.

@@

 \subsection{Estimate of mutual information and transfer entropy from relative frequencies}
 In order to compute the mutual information or the transfer entropy, we must estimate entropies from Eq.\ref{eq:entropy}.
 The histogram approach discretises the continuous sample space by partitioning it into bins that - in general - are not assumed to be of equal size, and so transforms the integral into a sum.
  \begin{eqnarray}
    \label{eq:entropyDiscrere}
    H(X) = - \int_{-\infty}^{+\infty}  {  p(x) \log{  p(x)}  } dx \simeq - \sum_{x_k}  \hat p(x_k) \log\left({ \hat p(x_k)}\right) v(x_k)\;.
  \end{eqnarray}
Where $\hat p(x_k)$ is the estimate of the probability density function $p(x)$ computed for the bin $x_k$ that contains the observation $x$.
The quantity $v(x_k)$ is the proportion of the support of the probability density function which is occupied by the bin. This generalises to multiple dimensions.
The estimate of the probability density function $\hat p(x_k)$ can be computed from the number of observations $n(x_k)$ that fall in each bin via
    \begin{eqnarray}
    \label{eq:pdfDiscrere}
 \hat p(x_k) = \frac{n(x_k)}{N v(x_k)}\;,
  \end{eqnarray}
where $N=\sum_{x_k} n(x_k)$ is the total number of observations. 
The entropy estimation is therefore:
  \begin{eqnarray}
    \label{eq:entropyDiscrere1}
    H(X) \simeq - \sum_{x_k}   \frac{n(x_k)}{N } \log\left({  \frac{n(x_k)}{N v(x_k)} }\right) \;.
  \end{eqnarray}
This quantity depends on the relative volumes of the bins, and hence do the estimates for $H(Y)$ and $H(X,Y)$ also require these to be calculated. However, in the mutual information $I(X;Y) = H(X)+H(Y)-H(X,Y)$ the terms containing the volumes cancel, and its expression turns out to be dependent only on the relative frequencies:
  \begin{eqnarray}
    \label{eq:mutualnfoDiscrere}
  I(X;Y) \! \simeq \!
 \! \!\sum_{x_k,y_j} \!\!\frac{n(x_k,y_j)}{N } \log\left({  \frac{n(x_k,y_j)}{N} }\right)
   -\! \sum_{x_k}      \! \frac{n(x_k)}{N } \log\left({  \frac{n(x_k)}{N} }\right) 
  - \!\sum_{y_j}        \!\frac{n(j_j)}{N } \log\left({  \frac{n(y_j)}{N} }\right) .
  \end{eqnarray}
  The same applies to the transfer entropy. 
  This is a remarkable simplification of the computation of these quantities, which shows how they can be estimated from histograms simply in terms of the relative frequencies, for any kind of binning partition.

@@

    \begin{tabular}{lrrrr}
      {} & {} & ({\color{blue} returns} $\rightarrow$ Sentiment) & {} & (Sentiment $\rightarrow$ {\color{blue} returns})  \\
      {} & Linear  &  NonLinear  &  Linear  &  Nonlinear  \\
      \hline 
      % BTC &     0.1 &     2.0 &        5.1 &        7.6 \\
      % LTC &     2.2 &    11.8 &       22.4 &       17.3 \\
      % XRP &     4.1 &     0.4 &       24.1 &       14.4 \\
      % ETH &    -0.6 &     0.5 &        6.5 &        6.8 \\
      BTC &     0.1 &       5.1 &       2.0 &        7.6 \\
      LTC &     2.2 &      22.4 &      11.8 &       17.3 \\
      XRP &     4.1 &      24.1 &       0.4 &       14.4 \\
      ETH &    -0.6 &       6.5 &       0.5 &        6.8 \\
      \end{tabular}

@@

%%%%%%%%%% Acknowledgements etc. %%%%%%%%%%%%%%

  \vskip10pc

  \dataccess{Social sentiment Data was provided courtesy of PsychSignal, and may be made available pending request to the authors.}

@@

  \ack{The authors acknowledge Th\'arsis Souza in advising on the method of testing for linear Granger causality, with thanks along with Yuqing Long, whose data collation and wrangling was a great help. Finally, a great debt of thanks is to PsychSignal for providing their market sentiment data for this academic study.}

>>>>

  In modern portfolio theory, investors commonly calculate correlations between asset types to construct portfolios aiming to maximise their return at a given level of risk \cite{markowitz1952portfolio}. In the search for excess returns, quantitative approaches are often exploited to detect predictive signals across time series. {\color{blue} This may be considered as a field of price discovery, which represents a large and fundamental area of financial theory.} In the ideal case {with complete predictive information, \color{blue}from} knowing the movement of one price, the movement of a second can be inferred {\color{blue} exactly. In reality information is incomplete and held asymmetrically, and the ability to reliably obtain this and contribute to price formation is generally rewarded by the market}. In this paper we explore the effectiveness of two promising techniques for detecting information transfer between time series, applying these to identify causal signals between alternative data and cryptocurrency {\color{blue} price returns}.

@@

 \subsection{Estimate{\color{blue}s} of Mutual Information and Transfer Entropy from Relative Frequencies}
 In order to compute the mutual information or the transfer entropy, we must estimate entropies from Eq.\ref{eq:entropy}.
 The histogram approach discretises the continuous sample space by partitioning it into bins that - in general - are not assumed to be of equal size, and so transforms the integral into a sum.
  \begin{eqnarray}
    \label{eq:entropyDiscrere}
    H(X) = - \int_{-\infty}^{+\infty}  {  p(x) \log{  p(x)}  } dx \simeq - \sum_{x_k}  \hat p(x_k) \log\left({ \hat p(x_k)}\right) v(x_k)\;.
  \end{eqnarray}
Where $\hat p(x_k)$ is the estimate of the probability density function $p(x)$ computed for the bin $x_k$ that contains the observation $x$.
The quantity $v(x_k)$ is the proportion of the support of the probability density function which is occupied by the bin. This generalises to multiple dimensions.
The estimate of the probability density function $\hat p(x_k)$ can be computed from the number of observations $n(x_k)$ that fall in each bin via
    \begin{eqnarray}
    \label{eq:pdfDiscrere}
 \hat p(x_k) = \frac{n(x_k)}{N v(x_k)}\;,
  \end{eqnarray}
where $N=\sum_{x_k} n(x_k)$ is the total number of observations. 
The entropy estimation is therefore:
  \begin{eqnarray}
    \label{eq:entropyDiscrere1}
    H(X) \simeq - \sum_{x_k}   \frac{n(x_k)}{N } \log\left({  \frac{n(x_k)}{N v(x_k)} }\right) \;.
  \end{eqnarray}
This quantity depends on the relative volumes of the bins, and hence do the estimates for $H(Y)$ and $H(X,Y)$ also require these to be calculated. However, in the mutual information $I(X;Y) = H(X)+H(Y)-H(X,Y)$ the terms containing the volumes cancel, and its expression turns out to be dependent only on the relative frequencies:
  \begin{eqnarray}
    \label{eq:mutualnfoDiscrere}
  I(X;Y) \! \simeq \!
 \! \!\sum_{x_k,y_j} \!\!\frac{n(x_k,y_j)}{N } \log\left({  \frac{n(x_k,y_j)}{N} }\right)
   -\! \sum_{x_k}      \! \frac{n(x_k)}{N } \log\left({  \frac{n(x_k)}{N} }\right) 
  - \!\sum_{y_j}        \!\frac{n(j_j)}{N } \log\left({  \frac{n(y_j)}{N} }\right) .
  \end{eqnarray}
  The same applies to the transfer entropy. 
  This is a remarkable simplification of the computation of these quantities, which shows how they can be estimated from histograms simply in terms of the relative frequencies, for any kind of binning partition.

@@

    \begin{tabular}{lrrrr}
      {} & {} & ({\color{blue} Returns} $\rightarrow$ Sentiment) & {} & (Sentiment $\rightarrow$ {\color{blue} Returns})  \\
      {} & Linear  &  NonLinear  &  Linear  &  Nonlinear  \\
      \hline

      BTC &     0.1 &       5.1 &       2.0 &        7.6 \\
      LTC &     2.2 &      22.4 &      11.8 &       17.3 \\
      XRP &     4.1 &      24.1 &       0.4 &       14.4 \\
      ETH &    -0.6 &       6.5 &       0.5 &        6.8 \\
      \end{tabular}

@@

%%%%%%%%%% Acknowledgements etc. %%%%%%%%%%%%%%
  \clearpage
  \vskip10pc

  \ethics{In this manuscript, there are no data or topics that need ethical approval.}

  \dataccess{The source code and data can be found at \url(https://github.com/ZacKeskin/). Synthetic data can be generated using the functions in the PyCausality package within test utils, in Time\_Series\_Generate.py.}

  \competing{There are no competing interests.}

  \aucontribute{ZK wrote the code, and drafted the first version of the paper. TA directed the research proposing the methodologies and applications. Both authors contributed equally to the final draft of the paper.}

  \ack{The authors acknowledge Th\'arsis Souza in advising on the method of testing for linear Granger causality, with thanks along with Yuqing Long, whose data collation and wrangling was a great help. Finally, we thank PsychSignal for providing their market sentiment data for this academic study.}

@@
